Bias in the Estimation of the Value of VRA over URA: A case of Mis-speficied Geographically Weighted Regression Models

### 2023-11-11 Writing/manuscript_gwr.rmd
rewrote the introduction

Precision agriculture technology allows input management strategies to adapt to the spatial and temporal variability of soil and crop conditions within a field. Over the past few decades, numerous technological innovations have been made for collecting detailed field information (e.g., yield mapping, soil sampling, remote sensing, etc.) and in developing site-specific management recommendations based on the information. But the long-standing question of whether the economic benefit of precision agriculture can offset the technology and information costs [@lowenberg1996revolution; @weiss1996precision] remains difficult to answer, particularly in the context of large-scale production systems [@mcfadden2022information].

While as-applied yield and input data are relatively easy to collect by modern farming machinery, gathering detailed soil characteristics data is often costly. On any given field, it is typical for many row crop farmers to pay for 1-ha (2.5-ac) grid soil sampling and chemical analysis every three or four years [@morris2018strengths]. The high expense of soil sampling and analysis is also driving development of new sensor technologies that gather data on site-specific field and crop characteristics or proxies of those characteristics, for example, mounted optical sensors for vegetative index (Trimble Agriculture's GreenSeeker and AgLeader's OptRx), towed soil sensor for electroconductivity data (Veris Technologies), etc. Those data are then converted into recommended input application rate maps (usually called "prescriptions" or "$R_x$s") by applying formulas provided by agricultural companies (e.g., FieldRevealTM, FieldApex) or university extension services (e.g., @culman2020tri). But deploying sophisticated sensors to gather field characteristics data is also costly, and the effects of their generation and use on farming profitability remains in question [@gardner2021economic].

In recent years, on-farm precision experimentation (OFPE) has gained traction with agricultural researchers and practitioners as a means of generating inexpensive, high-resolution yield and input trial data to inform site-specific crop input management [@bullock2019data; @lacoste2021farm]. Accompanying interest is surging among researchers in identifying efficient statistical methods to estimate site-specific yield response functions with OFPE data. Researchers recently have examined various methods, ranging from spatial econometrics [@anselin2004spatial; @liu2006site], Geographically Weighted Regression (GWR) [@rakshit2020; @evans2020assessment; @trevisan2021spatial] to machine learning techniques (suhc as random forest [@krause2020random; @de2023predicting], convolutional neural network [@barbosa2020modeling], and causal forest [@kakimoto2022causal]). Among these mehtods, however, most still require detailed soil sampling or field characteristics data together with the OFPE (yield, input) data for the site-specific estimation. The GWR model, on the other hand, runs a localized regression by taking only (yield, input) data from neighboring points around a site of interest, where the locality is controlled by the bandwidth of the kernel function chosen by the researcher. Therefore, it can estimate site-specific yield response functions without the costly soil and field characteristics data, and is appealing in OFPE data analysis.

@evans2020assessment explored using GWR with a linear model specification to detect low- and high-response areas, but like @rakshit2020, did not estimate site-specific economically optimal nitrogen rates.

@rakshit2020 also estimated site-specific yield response functions by applying GWR to on-farm experiment data assuming a quadratic functional form of yield response.  But they did not calculate site-specific economically optimal nitrogen rates (EONRs), and so unlike @trevisan2021spatial, did not conduct economic analysis.

Using OFPE data, @trevisan2021spatial assumed a quadratic yield response function, and reported large economic benefits associated with using GWR to generate site-specific seed and nitrogen rates recommendations.

@trevisan2021spatial demonstrated a method of providing site-specific management recommendations without such data. They estimated site-specific yield response functions by applying GWR techniques on OFPE data, and reported implausibly large results, estimating an average potential gain of \$65 ha$^{-1}$ from applying seed and nitrogen (N) at optimal uniform rates instead of at the rates the farmers would have chosen if not participating in the research, and an additional potential gain of \$58 ha$^{-1}$ when changing from the optimal uniform application rate to optimal site-specific application rates.

This might be seen as surprising given the often-reported poor statistical accuracy of GWR estimates [@griffith2008spatial; @harris2019simulation; @paez2011simulation; @da2016multiple] and also the much lower value of variable rate application (VRA) management over uniform rate application (URA) management reported in previous studies [@bullock1998does; @koch2004economic; @thrikawala1999economic].

* The central interest is the economic performance of GWR. We measure it by the profit improvement of GWR-based VRA over URA. We do not focus on statistical performance, as that is not the 
final goal

* Why we use simulation

## Sources of Bias in Estimating the Value of VRA with Mis-specified Yield Response

There are two potentially problematic factors that may have caused significant over-estimation of the value of GWR as a tool to estimate site-specific economically optimal nitrogen application rates in @trevisan2021spatial. First, @trevisan2021spatial assumed a quadratic functional form in their GWR estimation even though there is a wide consensus among the scientific community that corn yield response to nitrogen is better represented by the quadratic-plateau functional form. It is likely that their choice of quadratic functional form was merely a compromise because no statistical software package implements GWR with a quadratic-plateau functional form. The second source of over-estimation bias is endemic to any study that evaluates the value of variable rate application. In general, researchers take the following steps to estimate the value of VRA using OFPE data [@anselin2004spatial; @bullock2002adding; @liu2006site; @swinton1998evaluating]: (1) estimate site-specific yield response functions based on a method chosen by the researcher; (2) use the estimated yield response functions to estimate the economically optimal site-specific input rates (EOIRs) and economically optimal uniform input rates; (3) plug the estimated site-specific EOIRs and uniform EOIR into the estimated yield response functions to calculate estimated profits under both strategies; and (4) find the difference in the two strategies' estimated profits. However, this approach can result in significant bias in the estimation of the value of VRA because of step 3's assumption that the estimated yield response functions are the true yield response functions. This assumption is generally unavoidable for researchers since they cannot observe any real field's true site-specific yield response functions. But inaccurate estimation of yield response function in this methodology can result in significant bias in the estimation of the value of VRA. This bias problem can be particularly severe when the method used tends to exaggerate the true heterogeneity in yield response functions, such as GWR [@czarnota2015evaluating]. We hypothesize that the high value of VRA estimated by @trevisan2021spatial displays this bias.

We hypothesize that the economic value of GWR reported by @trevisan2021spatial is a result of over-estimation bias. There are two key sources of bias in their approach. First, they assumed a quadratic functional form of yield response, even though a consensus has evolved in the academic literature that the quadratic-plateau functional form better represents the yield-nitrogen relationship. Second, as will be discussed in detail in the following section, their approach suffers from bias that is endemic to any study that tries to estimate the value of a statistical method in the OFPE context, which is that it is necessary because the true yield response function is unknowable and can only be estimated.

The primary objective of this study is to examine the degree and direction of bias in the estimation of the economic value of GWR-based site-specific nitrogen rate recommendations (i.e., variable rate applications of nitrogen) when the yield response function assumed in the GWR model is specified as quadratic but is actually of quadratic-plateau functional form.  We will do so using Monte Carlo simulations, which, since we define and know the true yield response function in the data generating process, allow us to measure the degree of bias. Unlike in @griffith2008spatial, @harris2019simulation, @paez2011simulation, and @da2016multiple, our objective is not to measure the bias in the coefficient estimation by GWR, but rather to measure economic effects^[While the environmental impacts of fertilizer management are of great social importance, it is out of the scope of this study. The "economically optimal" in the context of this study does not mean "socially optimal", but is rather limited to the economic optimization of farmers' private production.] of the statistical inaccuracy in estimating the coefficients of the model.  We show that the misspecification tends to significantly over-estimate GWR's true economic value.

## Conceptual Framework: Estimation of Yield-to-input Response Relationships Using OFPE Data
OFPE can provide detailed as-applied input and yield data at dense resolutions over entire fields, which if well analyzed may provide valuable information that can improve site-specific input application recommendations [@bullock2019data; @lacoste2021farm]. Consider a yield response function $y = f(\textbf{x}, \textbf{c})$, where following @bullock2000agronomic, $\textbf{x}$ is a vector of input management variables and $\textbf{c}$ is a vector of spatial variables describing field and soil characteristics^[For simplicity, we suppress discussion of the need to bring variance in weather data into yield response estimation.  Obviously, generation and use of weather data is necessary to provide meaningful probabilistic analysis and input management advice.]. Letting a field site $i$ be defined as a contiguous area characterized uniformly by a value $\textbf{c}_i$ of the vector of field and soil characteristics, its *site-specific yield response function* is the reduced form of the field's yield response function created when the value of the vector of characteristics variables is set at the site's value, $\textbf{c}_i$:  $f_i(\textbf{x}) \equiv f(\textbf{x}, \textbf{c}_i)$.  This conceptualization suggests two methods of estimating a field's site-specific yield responses. The first is to estimate a "map" of the field characteristics, for example thought of as a set of the field's $n \geq 1$ sites' values of the characteristics vector: $\{\textbf{c}_1,\dots, \textbf{c}_n\}$.  With that map, regressing yield on $\textbf{x}$ and $\textbf{c}$ generates an estimated "field-specific" yield response function $\hat{f}(\textbf{x},\textbf{c})$. Estimates of site-specific yield response functions can then be obtained by inserting the value of the site's characteristic variables into the estimate of the field-specific response function: $\hat{f}_i(\textbf{x}) \equiv \hat{f}(\textbf{x},\textbf{c}_i)$. Alternatively, a site-specific yield response function might be estimated simply by taking (input rate, yield) data from points around a site of interest $i$, and regressing yield on input application rates to attain the estimate $\hat{f}_i (\textbf{x})$. Stated roughly, using GWR takes the latter approach, which has the advantage of not needing a costly map of field characteristics values.

@@

>>>>

Precision agriculture technology allows input management strategies to adapt to the spatial and temporal variability of soil and crop conditions within a field. Over the past few decades, numerous technological innovations have been made for collecting detailed field information (e.g., yield mapping, soil sampling, remote sensing, etc.) and in developing site-specific management recommendations based on the information. In recent years, on-farm precision experimentation (OFPE) has gained traction with agricultural researchers and practitioners as a means of generating inexpensive, high-resolution yield and input trial data to inform site-specific crop input management [@bullock2019data; @lacoste2021farm].  One of the crucial and long-stanging research question is how valuable variable rate application (VRA) of input (e.g., fertilizer, seed) over the traditional uniform rate application (URA). This is because the answer to this question determines whether farmers should invest in expensive technologies, conduct soil sampling, or purchase third-party variable rate prescription map.

Unforuntealy, the estimation of the value of VRA over URA is inherently biased. In general, researchers take the following steps to estimate the value of VRA using OFPE data [@anselin2004spatial; @bullock2002adding; @liu2006site; @swinton1998evaluating]: (1) estimate site-specific yield response functions based on a method chosen by the researcher; (2) use the estimated yield response functions to estimate the economically optimal site-specific input rates (EOIRs) and economically optimal uniform input rates; (3) plug the estimated site-specific EOIRs and uniform EOIR into the estimated yield response functions to calculate estimated profits under both strategies; and (4) find the difference in the two strategies' estimated profits. The core source of bias lies in step (3). It implicitly assumes that the estimated site-specific yield response functions are true. Unfortunately, this assumption is not avoidable in general since we will never know the true site-specific yield response functions. Consequently, when the estimated site-specific yield response functions are inaccurate and significanly different from the true site-specific yield response functions, it is not hard to imagine that estimation of the value of VRA over URA is also biased. The direction of bias can be either negative or positive. One way to exercebate the bias is functional form mis-specificaiton because it can immediately introduce mis-reppresentation of the true yield response functions.

There are mainly two strands of approaches in obtaining VRA recommendations using OFPE data. First strand uses soil sampling and/or field characteristics (e.g., topography) data together with the OFPE (yield, input) data to estimate site-specific yield response functions. Many different statistical methods have been used under this approach, ranging from spatial econometrics [@anselin2004spatial; @liu2006site] to machine learning methods, suhc as random forest [@krause2020random; @de2023predicting], convolutional neural network [@barbosa2020modeling], and causal forest [@kakimoto2022causal]. The second strand is a characteristic-agnostic approach, represented by geographically weighted regression (GWR), which runs a localized regression by taking only yield- input data from neighboring points around a site of interest [@rakshit2020; @evans2020assessment; @trevisan2021spatial]. Therefore, it can estimate site-specific yield response functions without site-specific characteristics. The benefit of this approach is that it does not require soil sampling or sensing, which can be highly costly^[On any given field, it is typical for many row crop farmers to pay for 1-ha (2.5-ac) grid soil sampling and chemical analysis every three or four years [@morris2018strengths]. The high expense of soil sampling and analysis is also driving development of new sensor technologies that gather data on site-specific field and crop characteristics or proxies of those characteristics, for example, mounted optical sensors for vegetative index (Trimble Agriculture's GreenSeeker and AgLeader's OptRx), towed soil sensor for electroconductivity data (Veris Technologies), etc. Those data are then converted into recommended input application rate maps (usually called "prescriptions" or "$R_x$s") by applying formulas provided by agricultural companies (e.g., FieldRevealTM, FieldApex) or university extension services (e.g., @culman2020tri). But deploying sophisticated sensors to gather field characteristics data is also costly, and the effects of their generation and use on farming profitability remains in question [@gardner2021economic]].

Among the GWR studies, @trevisan2021spatial estimated site-specific yield response functions by applying GWR on OFPE data, and reported implausibly large estimate of the potential gain of \$58 ha$^{-1}$ when changing from the optimal uniform application rate to optimal site-specific application rates. This is surprising given the much lower value of variable rate application (VRA) management over uniform rate application (URA) management reported in previous studies [@bullock1998does; @koch2004economic; @thrikawala1999economic]. We hypothesize that their estimate of the value of GWR-based VRA over URA are significantly positively biased. First, they most likely mis-specified the model of corn yield reponse to nitrogen. It has been widely shown in agronomic studies that corn yield exhibits quadratic-plaeau response. However, they used quadratic model. GWR has also been criticized for its poor statistical accuracy of coefficient estimation [@griffith2008spatial; @harris2019simulation; @paez2011simulation; @da2016multiple]. We hypothesize that the combination of these two factors may have results in significant over-estimation of the value of VRA over URA.

The primary objective of this study is to examine the degree and direction of bias in the estimation of the economic value of GWR-based site-specific nitrogen rate recommendations (i.e., variable rate applications of nitrogen) when the yield response function assumed in the GWR model is specified as quadratic but is actually of quadratic-plateau functional form.  We will do so using Monte Carlo simulations, which, since we define and know the true yield response function in the data generating process, allow us to measure the degree of bias. Unlike in @griffith2008spatial, @harris2019simulation, @paez2011simulation, and @da2016multiple, our objective is not to measure the bias in the coefficient estimation by GWR, but rather to measure economic effects^[While the environmental impacts of fertilizer management are of great social importance, it is out of the scope of this study. The "economically optimal" in the context of this study does not mean "socially optimal", but is rather limited to the economic optimization of farmers' private production.] of the statistical inaccuracy in estimating the coefficients of the model. We show that the mis-specified GWR indeed tends to significantly over-estimate GWR's true economic value.

@@

# Conceptual Framework: Estimation of Yield-to-input Response Relationships Using OFPE Data

OFPE can provide detailed as-applied input and yield data at dense resolutions over entire fields, which if well analyzed may provide valuable information that can improve site-specific input application recommendations [@bullock2019data; @lacoste2021farm]. Consider a yield response function $y = f(\textbf{x}, \textbf{c})$, where following @bullock2000agronomic, $\textbf{x}$ is a vector of input management variables and $\textbf{c}$ is a vector of spatial variables describing field and soil characteristics^[For simplicity, we suppress discussion of the need to bring variance in weather data into yield response estimation.  Obviously, generation and use of weather data is necessary to provide meaningful probabilistic analysis and input management advice.]. Letting a field site $i$ be defined as a contiguous area characterized uniformly by a value $\textbf{c}_i$ of the vector of field and soil characteristics, its *site-specific yield response function* is the reduced form of the field's yield response function created when the value of the vector of characteristics variables is set at the site's value, $\textbf{c}_i$:  $f_i(\textbf{x}) \equiv f(\textbf{x}, \textbf{c}_i)$.  This conceptualization suggests two methods of estimating a field's site-specific yield responses. The first is to estimate a "map" of the field characteristics, for example thought of as a set of the field's $n \geq 1$ sites' values of the characteristics vector: $\{\textbf{c}_1,\dots, \textbf{c}_n\}$.  With that map, regressing yield on $\textbf{x}$ and $\textbf{c}$ generates an estimated "field-specific" yield response function $\hat{f}(\textbf{x},\textbf{c})$. Estimates of site-specific yield response functions can then be obtained by inserting the value of the site's characteristic variables into the estimate of the field-specific response function: $\hat{f}_i(\textbf{x}) \equiv \hat{f}(\textbf{x},\textbf{c}_i)$. Alternatively, a site-specific yield response function might be estimated simply by taking (input rate, yield) data from points around a site of interest $i$, and regressing yield on input application rates to attain the estimate $\hat{f}_i (\textbf{x})$. Stated roughly, using GWR takes the latter approach, which has the advantage of not needing a costly map of field characteristics values.
